---
layout: post
title: "Alpha Shape Complexes"
author: "Ben Holmgren"
date: "December 12, 2019"
---

{% include lib/mathjax.html %}


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Alpha Shapes and Alpha Shape Complexes

### Objectives
* Gain background knowledge of Delaunay triangulation in order to conceptualize the definition of an alpha complex.
* Understand the difference between an alpha shape and an alpha complex.
* Gain insight into the usefulness of Alpha shapes and complexes in persistent homology.
* Be able to apply Alpha shapes and Alpha complexes to persistent homology.

### Background
In order to understand alpha shapes and alpha complexes, it is first useful to have some background knowledge pertaining to Delaunay Triangulation.
Given a set of discrete points in a plane we can triangulate points such that no point lies inside the circumcircle of a triangle. What does this mean?
Envision a set of connected points:

```{r points, echo=FALSE}
plot(0:10, 0:10, type = "n")
x <- c(1,3,6,7,2,3,4)
y <- c(4,5,6,3,9,10,1)
lines(c(1,3),c(4,5))
lines(c(1,2),c(4,9))

#TODO: figure out how to run circumcircle
#circumcircle(c(1,3,6),c(4,5,6), plot = TRUE)

vs <- c(c(1,4),c(3,5),c(6,6),c(7,3),c(2,9),c(3,10),c(4,1),nrow = 2,ncol = 7)
for(i in 1:7){
  points(x[i],y[i],pch=20)
}

```

## Definition
Alpha complexes then, are a simplicial complex which is contained in the Delaunay triangulation. We find our alpha complex by inserting balls around each vertex, and adding edges with the intersection of balls between corresponding vertices. 
For example:
#TODO:make plot here illustrating this

.
